Finiteness in Cubical Type Theory

### 2020-06-25 sections/search.tex
Update search.tex

We have already, in previous sections, explored the theoretical implications of
Cubical Type Theory on our formalisation.
With this library for proof search, however, we will see two distinct
\emph{practical} applications which would simply not be possible without
computational univalence.
First and foremost: our proofs of finiteness, constructed with the API we will
describe, have all the power of full equalities.
Put another way any proof over a finite type \(A\) can be lifted to any other
type with the same cardinality.
Secondly our proof search can range over functions: we could, for instance, have
asked the prover to find if \emph{any} function over \(\AgdaDatatype{Bool}\) is
associative, and if so return it to us.
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{some-assoc}
\end{agdalisting}
The usefulness of which is dubious, but we will see a more interesting
application soon.
\section{Proof Automation And Search Techniques}
For this prover we will not resort to reflection or similar techniques: instead,
we will trick the type checker to do our automation for us.
This is a relatively common technique, although not so much outside of Agda, so
we will briefly explain it here.

@@

So our task is to rewrite proof obligations like the one in
Equation~\ref{bool-assoc-auto-proof} into ones which can reduce completely. 
As it turns out, we have already described the type of proofs which can ``reduce
completely'': \emph{decidable} proofs.
If we have a decision procedure over some proposition \(P\) we can run that
decision during type checking, because the decision procedure itself is a proof
that the decision will terminate.
In code, we capture this idea with the following pair of functions:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{is-true}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{from-true}
  \end{agdalisting}
\end{multicols}
The first is a function which derives a type from whether a decision is
successful or not.
This function is important because if we use the output of this type at any
point we will effectively force the unifier to run the decision computation.
The second takes---as an implicit argument---an inhabitant of the type generated
from the first, and uses it to prove that the decision can only be true, and the
extracts the resulting proof from that decision.
All in all, we can use it like this:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{extremely-obvious}
\end{agdalisting}
This technique will allow us to automatically compute any decidable predicate.
\section{Omniscience}
So we now know what is needed of us for proof automation: we need to take our
proofs and make them decidable.
In particular, we need to be able to ``lift'' decidability back over a
function arrow.
For instance, given \(x\), \(y\), and \(z\) we already have
\(\AgdaDatatype{Dec}\;((x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z))\)
(because equality over booleans is decidable).
In order to turn this into a proof that \AgdaFunction{∧} is associative we need
\(\AgdaDatatype{Dec}\;(\forall \; x \; y \; z \rightarrow (x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z))\).
The ability to do this is described formally by the notion of
``Exhaustibility''.
\begin{definition}[Exhaustibility]
  We say a type \(A\) is exhaustible if, for any decidable predicate \(P\) on
  \(A\), the universal quantification of the predicate is decidable.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Omniscience.tex]{exhaustible}
  \end{agdalisting}
\end{definition}
This property of \AgdaDatatype{Bool} would allow us to automate the proof of
associativity, but it is in fact not strong enough to find individual
representatives of a type which support some property.
For that we need the more well-known, but related, property of
\emph{omniscience}.
\begin{definition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience \cite{myhillErrettBishopFoundations1972} is as follows:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Omniscience.tex]{omniscient}
  \end{agdalisting}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{definition}
Both of these properties only make sense to talk about in a constructive
setting: classically speaking, \emph{every} type is omniscient and exhaustible.
Finite types are omniscient (the reason should be clear, but we will prove this
fact presently), but perhaps surprisingly, it is not \emph{only} finite types
which are exhaustible.
Certain infinite types can be
exhaustible~\cite{escardoInfiniteSetsThat2007}, but an exploration of that is
beyond the scope of this work.

Omniscience and exhaustibility are not interchangeable: conceptually,
omniscience needs some kind of ordering on the underlying type.
If we had a property, say like the associativity property above, that is
satisfied by more than one element of the type, we would have to have some way
to pick a candidate element from the type to return in the case of the property
being satisfied.
This requirement of order mirrors the Ord difference in
Fig.~\ref{finite-classification}.

Omniscience implies exhaustibility, then, but the converse is not true.
\begin{lemma}
  Omniscience implies exhaustibility
\end{lemma}
\begin{proof}
  \todo{proof here}
\end{proof}

>>>>

We have already, in previous sections, explored the theoretical implications of
Cubical Type Theory on our formalisation.
With this library for proof search, however, we will see two distinct
practical applications which would simply not be possible without
computational univalence.
First and foremost: our proofs of finiteness, constructed with the API we will
describe, have all the power of full equalities.
Put another way any proof over a finite type \(A\) can be lifted to any other
type with the same cardinality.
Secondly our proof search can range over functions: we could, for instance, have
asked the prover to find if \emph{any} function over \(\AgdaDatatype{Bool}\) is
associative, and if so return it to us.
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{some-assoc}
\end{agdalisting}
The usefulness of which is dubious, but we will see a more interesting
application soon.
\section{Proof Automation And Search Techniques}
For this prover we will not resort to reflection or similar techniques: instead,
we will trick the type checker to do our automation for us.
This is a relatively common technique, although not so much outside of Agda, so
we will briefly explain it here.

@@

So our task is to rewrite proof obligations like the one in
Equation~\ref{bool-assoc-auto-proof} into ones which can reduce completely. 
As it turns out, we have already described the type of proofs which can ``reduce
completely'': \emph{decidable} proofs.
If we have a decision procedure over some proposition \(P\) we can run that
decision during type checking, because the decision procedure itself is a proof
that the decision will terminate.
In code, we capture this idea with the following pair of functions:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{is-true}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{from-true}
  \end{agdalisting}
\end{multicols}
The first is a function which derives a type from whether a decision is
successful or not.
This function is important because if we use the output of this type at any
point we will effectively force the unifier to run the decision computation.
The second takes---as an implicit argument---an inhabitant of the type generated
from the first, and uses it to prove that the decision can only be true, and the
extracts the resulting proof from that decision.
All in all, we can use it like this:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{extremely-obvious}
\end{agdalisting}
This technique will allow us to automatically compute any decidable predicate.
\section{Omniscience}
So we now know what is needed of us for proof automation: we need to take our
proofs and make them decidable.
In particular, we need to be able to ``lift'' decidability back over a
function arrow.
For instance, given \(x\), \(y\), and \(z\) we already have
\(\AgdaDatatype{Dec}\;((x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z))\)
(because equality over booleans is decidable).
In order to turn this into a proof that \AgdaFunction{∧} is associative we need
\(\AgdaDatatype{Dec}\;(\forall \; x \; y \; z \rightarrow (x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z))\).
The ability to do this is described formally by the notion of
``Exhaustibility''.
\begin{definition}[Exhaustibility]
  We say a type \(A\) is exhaustible if, for any decidable predicate \(P\) on
  \(A\), the universal quantification of the predicate is decidable.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Omniscience.tex]{exhaustible}
  \end{agdalisting}
\end{definition}
This property of \AgdaDatatype{Bool} would allow us to automate the proof of
associativity, but it is in fact not strong enough to find individual
representatives of a type which support some property.
For that we need the more well-known, but related, property of
\emph{omniscience}.
\begin{definition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience \cite{myhillErrettBishopFoundations1972} is as follows:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Omniscience.tex]{omniscient}
  \end{agdalisting}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{definition}

These properties are closely related to the axiom of choice: in particular, the
statement that ``every type is omniscient'' implies the AOC. \todo{Reference for
this, and possible proof? Also it might not be entirely correct: I think you
need a truncation somewhere for omniscience to be equivalent to the AOC}
Because we're constructive, only a select few types are omniscient: finite
types, for instance (why finite types are omniscient is not too difficult to
see, but we will explain it soon).
Perhaps surprisingly, it is not \emph{only} finite types
which are exhaustible.
Certain infinite types can be
exhaustible~\cite{escardoInfiniteSetsThat2007}, but an exploration of that is
beyond the scope of this work.

Omniscience and exhaustibility are not interchangeable: every omniscient type is
exhaustible, but the converse is not true.
Conceptually, omniscience needs some kind of ordering on the underlying type.
This is because omniscience returns a candidate satisfying the given predicate:
there is no requirement, though, that only one element in the underlying type
satisfies the decidable predicate.
As a result, omniscience needs some way to choose among all possible candidates:
this is the ``order'' we are referring to.
This is also the same ``order'' that we referred to when talking about the
finiteness predicates: all ordered predicates (in
Figure~\ref{finite-classification}) imply omniscience, whereas the unordered
predicates only imply exhaustibility.
\begin{lemma}
  Omniscience implies exhaustibility
\end{lemma}
\begin{proof}
  \todo{proof here}
\end{proof}
